---
title: "Exercise 2"
author: "Jon Guler"
date: "2024-04-26"
output: html_document
---

```{r}
library("readr")
library("ggplot2")
library("sf")
library("dplyr")
library("terra")
library("tmap")
library(tidyr)
```

#Demo

```{r Demo}
now <- as.POSIXct("2024-04-26 10:22:00")
later <- as.POSIXct("2024-04-26 11:35:00")
time_difference <- as.numeric(difftime(later, now, units = "secs"))

difftime_secs <- function(later, now){
  as.numeric(difftime(later, now, units = "secs"))
}

numbers <- 1:10
lead(numbers, 2)
lag(numbers, 2, default = 9999)

wildschwein <- tibble(
    TierID = c(rep("Hans", 5), rep("Klara", 5)),
    DatetimeUTC = rep(as.POSIXct("2015-01-01 00:00:00", tz = "UTC") + 0:4 * 15 * 60, 2)
)

now2 <- wildschwein$DatetimeUTC
later2 <- lead(now2)
wildschwein$timelag <- difftime_secs(later2, now2)

wildschwein <- group_by(wildschwein, TierID)
mutate(wildschwein,
       timelag2 = difftime_secs(lead(DatetimeUTC), DatetimeUTC)
       )

wildschwein <- wildschwein %>%
  group_by(TierID) %>%
  mutate(
    timelag2 = difftime_secs(lead(DatetimeUTC), DatetimeUTC)
  )

wildschwein <- wildschwein %>%
  group_by(TierID) %>%
  summarise(
    first_sample = min(DatetimeUTC)
  )

```

#Exercise A

##Task 1

```{r Task 1.1}
wildschwein_BE <- read_delim("Data/wildschwein_BE_2056.csv", ",")
wildschwein_BE <- st_as_sf(wildschwein_BE, coords = c("E", "N"), crs = 2056)
```

##Task 2

```{r Task 1.2}
difftime_secs <- function(later, now){
    as.numeric(difftime(later, now, units = "secs"))
}

wildschwein_BE <- wildschwein_BE %>% 
  arrange(DatetimeUTC) %>%  # Ensure data is sorted by time
  mutate(time_lag = lead(DatetimeUTC) %>% difftime_secs(now = DatetimeUTC))

#Question 1
num_individuals <- length(unique(wildschwein_BE$TierID))

#Question 2
individual_tracking_summary <- wildschwein_BE %>%
  group_by(TierID) %>%
  summarise(
    start_date = min(DatetimeUTC),
    end_date = max(DatetimeUTC),
    duration = difftime(end_date, start_date, units = "days"),
    recorded_points = n(),
    avg_time_lag = mean(time_lag, na.rm = TRUE)
  )
```
Question 1:
Three individuals were tracked

Question 2:
- 002A: 338 days
- 016A: 234 days
- 018A: 261 days
Yes, there are time gaps of sometimes several hours

Question 3:
It began with a separate tracking but in the end, ll the individuals were tracked concurrently

Question 4:
The average time lag between two samples is the following:
- 002A: 603
- 016A: 434
- 018A: 657
Approximately all 15 minutes

##Task 3

```{r Task 1.3}
# Custom function to calculate distance between elements
distance_by_element <- function(later, now) {
  as.numeric(st_distance(later, now, by_element = TRUE))
}

# Calculate the lagged geometry to compare current and previous points
wildschwein_BE$geometry_lag <- lag(wildschwein_BE$geometry)

# Apply the distance function to create a new column 'step_length'
wildschwein_BE <- wildschwein_BE %>%
  mutate(steplength = distance_by_element(geometry_lag, geometry))
```

##Task 4

```{r Task 1.4}
# Define the function to calculate distance by elements
distance_by_element <- function(later, now) {
  as.numeric(st_distance(later, now, by_element = TRUE))
}

# Calculate step length by applying the distance function
wildschwein_BE <- wildschwein_BE %>%
  mutate(steplength = distance_by_element(geometry_lag, geometry))

# Calculate speed in meters per second (m/s)
wildschwein_BE <- wildschwein_BE %>%
  mutate(speed = steplength / time_lag)  # Assuming time_lag is in seconds and steplength in meters

# Clean up by removing the temporary geometry_lag column
wildschwein_BE$geometry_lag <- NULL
```

##Task 5

```{r Task 1.5}
wildschwein_sample <- wildschwein_BE %>%
  filter(TierName == "Sabi") %>% 
  head(100)

tmap_mode("view")

tm_shape(wildschwein_sample) + 
  tm_dots()

wildschwein_sample_line <- wildschwein_sample |> 
  # dissolve to a MULTIPOINT:
  summarise(do_union = FALSE) |> 
  st_cast("LINESTRING")

tmap_options(basemaps = "OpenStreetMap")

tm_shape(wildschwein_sample_line) +
  tm_lines() +
  tm_shape(wildschwein_sample) + 
  tm_dots()
```

#Exercise B

##Task 1

```{r Task 2.1}
caro <- read_delim("Data/caro60.csv", ",") |>
  st_as_sf(coords = c("E","N"), crs = 2056) |> 
  select(DatetimeUTC)

# Ensure data is sorted by DatetimeUTC if not already
caro <- caro %>% 
  arrange(DatetimeUTC)

# Calculate previous and next geometries
caro$geometry_prev <- lag(caro$geometry)
caro$geometry_next <- lead(caro$geometry)

# Calculate step length using midpoint (previous to next)
caro <- caro %>%
  mutate(steplength = distance_by_element(geometry_prev, geometry_next),
         timelag = 120,  # Since the sampling interval is fixed at 120 seconds
         speed = steplength / timelag)  # Speed in meters per second

# Remove the temporary geometry columns
caro <- select(caro, -c(geometry_prev, geometry_next))

# Check the first few rows to verify
print(head(caro))
```

##Task 2

```{r Task 2.2}
# Calculate the geometry with an offset of 2 using lead/lag functions
caro <- caro %>%
  mutate(geometry_plus_2 = lead(geometry, n = 2),  # Getting the geometry two steps ahead
         geometry_minus_2 = lag(geometry, n = 2))  # Getting the geometry two steps behind

# Apply the distance_by_element to calculate the step length with an offset of 2
caro <- caro %>%
  mutate(steplength2 = distance_by_element(geometry_minus_2, geometry_plus_2),
         timelag2 = 240,  # Time difference for 2 steps at 120 seconds each
         speed2 = steplength2 / timelag2)  # Calculate speed in meters per second

# Remove temporary geometry columns
caro <- select(caro, -c(geometry_plus_2, geometry_minus_2))

# Drop geometry and select only specific columns to display relevant data only
result <- caro %>%
  st_drop_geometry() %>%
  select(time2 = DatetimeUTC, timelag2, steplength2, speed2) %>%
  head()

# Display the first few rows to verify the calculations
print(result)
```

##Task 3

```{r Task 2.3}
# Calculate the geometry with an offset of 4 using lead/lag functions
caro <- caro %>%
  mutate(geometry_plus_4 = lead(geometry, n = 4),  # Getting the geometry four steps ahead
         geometry_minus_4 = lag(geometry, n = 4))  # Getting the geometry four steps behind

# Apply the distance_by_element to calculate the step length with an offset of 4
caro <- caro %>%
  mutate(steplength3 = distance_by_element(geometry_minus_4, geometry_plus_4),
         timelag3 = 480,  # Time difference for 4 steps at 120 seconds each
         speed3 = steplength3 / timelag3)  # Calculate speed in meters per second

# Remove temporary geometry columns
caro <- select(caro, -c(geometry_plus_4, geometry_minus_4))

# Drop geometry and select only specific columns to display relevant data only
result <- caro %>%
  st_drop_geometry() %>%
  select(time3 = DatetimeUTC, timelag3, steplength3, speed3) %>%
  head()

# Display the first few rows to verify the calculations
print(result)
```

##Task 4

```{r Task 2.4}
# before pivoting, let's simplify our data.frame
caro2 <- caro |> 
  st_drop_geometry() |> 
  select(DatetimeUTC, speed, speed2, speed3)

caro_long <- caro2 |> 
  pivot_longer(c(speed, speed2, speed3))
  
head(caro_long)

ggplot(caro_long, aes(name, value)) +
  # we remove outliers to increase legibility, analogue
  # Laube and Purves (2011)
  geom_boxplot(outliers = FALSE)
```

#Exercise C

##Task 1

```{r Task 3.1}
# Import GPX data
gpx_path <- "Data/Jogging 3.gpx"
gpx_data <- st_read(gpx_path, layer = "tracks")  # Change 'tracks' if a different layer is needed

# Transform the CRS to CH1903+ LV95 (EPSG:2056)
gpx_data_lv95 <- st_transform(gpx_data, crs = 2056)

# Plotting the data using ggplot2
ggplot(data = gpx_data_lv95) +
  geom_sf() +  # If points, use geom_sf(); if lines, ensure geom_sf() works as is
  theme_minimal() +
  labs(title = "Map of GPX Data",
       caption = "Data transformed to CH1903+ LV95")

# Create a map using tmap
tmap_mode("view")  # Use "plot" for static maps, "view" for interactive maps
tm_shape(gpx_data_lv95) +
  tm_lines() +  # tm_points() for point data
  tm_layout(title = "Map of GPX Data in CH1903+ LV95")
```
